Reducing Acceptance Marks \\ in Emerson-Lei Automata

### 2021-10-14 paper/main.tex
preliminaries polished

\def\Mocc{M_{\mathrm{occur}}}
\def\Mevery{M_{\mathrm{every}}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\CA}{\ensuremath{\mathcal{A}}}
\newcommand{\CB}{\ensuremath{\mathcal{B}}}
\def\init{\iota}
\def\lang{L} % FIX this
\newcommand{\cM}{\mathcal{M}}
\newcommand{\bscc}{\mathcal{B}}
\newcommand{\sched}{\mathfrak{s}}
\newcommand{\Act}{\mathit{Act}}
\newcommand{\NP}{\textsf{NP}}
\newcommand{\POLY}{\textsf{P}}
\newcommand{\class}{\ensuremath{\mathcal{C}}}
\newcommand{\finally}{\mathsf{F}}
\newcommand{\globally}{\mathsf{G}}
% \newcommand{\finally}{\mathbf{F}}
% \newcommand{\globally}{\mathbf{G}}

@@

\begin{abstract}
  The formalism called \emph{transition-based Emerson-Lei automata
    (TELA)} generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in TELA are marked with acceptance marks and
  accepting formulas are boolean combinations of terms saying that a
  particular mark has to be visited infinitely or finitely often in an
  accepting run. Algorithms processing these automata can be very
  sensitive to the number of different acceptance marks. We introduce
  two techniques reducing the number of acceptance marks in a TELA
  without altering its structure and the represented language. One
  technique represents a standard approach based on relations among
  acceptance marks in individual strongly connected components of the
  automaton. The other technique constructs \emph{quantified boolean
    formula (QBF)} queries that ask a QBF solver for an acceptance
  condition with fewer acceptance marks and the placement of these
  marks in the automaton structure. Both techniques are implemented
  and experiments show that the number of acceptance marks in TELA
  produced by state-of-the-art tools ltl3tela, Rabinizer~4, and SPOT
  can be sometimes reduced. In the case of Rabinizer~4, we reduced the
  cummulative number of acceptance marks to less than one half.
\end{abstract}

@@

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with traditional
acceptance conditions. Further, algorithms handling TELA can
automatically handle all automata with traditional acceptance
conditions. TELA automata can be obtained for example by translating
LTL formulas by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity automata~\cite{renkin.20.atva}.

@@

The second technique is completely different. For a given TELA and
parameters $c,m$, it produces a \emph{quantified boolean formula
  (QBF)} which is satisfiable if and only if there exists an automaton
with the same structure, $m$ acceptance marks, an acceptance formula
in disjunctive normal form with $c$ clauses, and the same set
of accepting runs. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of a
simpler formula such its satisfiability implies existence of an
automaton with the same structure, $m$ acceptance marks, and the same
set of accepting runs, but not vice versa.

@@

\begin{definition}[TELA]
  A \emph{transition-based Emerson-Lei automaton (TELA)} is a tuple
  $\mathcal{A} =(Q,M,\Sigma,q_{0},\delta,\varphi)$, where
  \begin{itemize}
  \item $Q$ is a finite set of \emph{states},
  \item $M$ is a finite set of \emph{acceptance marks},
  \item $\Sigma$ is a finite \emph{alphabet},
  \item $q_{0}\in Q$ is an \emph{initial state},
  \item $\delta \subseteq Q \times \Sigma \times 2^{M} \times Q$ is a
    \emph{transition relation}, and
  \item $\varphi$ is the \emph{acceptance condition} constructed by
    the following abstract syntax equation, where $m$ ranges over $M$:

    \begin{center}
        $\varphi ::= \true \mid \false \mid \Inf(m) \mid \Fin(m) \mid \left( \varphi \land \varphi \right) \mid \left( \varphi \lor \varphi \right)$.\\
    \end{center} 
\end{itemize}
\end{definition}

\todo{Honza je tu a jde spat (je 5 rano pryc).}

A run $\pi$ of $\mathcal{A}$ on word $u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence of transitions $\pi = (q_{0}, u_{0}, M_{0}, q_{1})(q_{2}, u_{1}, M_{1}, q_{2}) \dots \in \delta^{\omega}$.

Let $M(\pi)$ denote set of all acceptance marks, that occur in run $\pi$ in infinitely many transitions. The run satisfies $\Inf(m)$ $\mathit{iff} \ m \in M(\pi)$ and it satisfies $\Fin(m) \ \mathit{iff} \ m \ \not \in M(\pi)$.
Run $\pi$ is \emph{accepting} if it satisfies the formula $\varphi$. A set of words, which are accepted by $\mathcal{A}$, represent a language $\mathcal{L}(\mathcal{A})$.

Sometimes the set of transitions with the same acceptance mark $m\in M$ is referred to as \emph{acceptance set} $Z(m)$.

\begin{itemize}
    \item define DNF

\subsection{Strongly connected component}
Let us consider a TELA $\mathcal{A} = \left(  \mathcal{Q}, M, \Sigma,q_{0}, \delta, \varphi \right)$. \emph{Strongly connected component} (SCC) of $\mathcal{A}$ is each maximal set of states $S \subseteq Q$ such that any two states $q_{1}, q_{2} \in S$ are connected by a sequence of transitions.
Let us enumerate the SCCs of $\mathcal{A}$.
Set of all transitions of the SCC $S_{i} \text{ is } \delta_{S_{i}} = \delta \cap (S_{i} \times \Sigma \times 2^M \times S_{i})$. A run of SCC $S_{i}$ is a run of automaton $\mathcal{A}$ with the infinite suffix contained within $S_{i}$.

\subsection{Quantified Boolean formula}
\emph{Quantified Boolean formulae} (QBFs) is an extension of propositional logic -- some Boolean variables in the formula can be quantified with \emph{universal} or \emph{existential} quantifiers.
Formula $\Phi$ is a QBF formula in \emph{prenex normal form} $\iff$ $\Phi = Q_{1}x_{1}...Q_{n}x_{n} . \varphi$, where $Q_{i} \in \{\forall, \exists\}$, $x_{i}$ is a quantified Boolean variable and $\varphi$ is quantifier-free Boolean formula.

\smallskip 
Let $\Phi$ be a QBF. Semantically we define the \emph{universal} quantifier as $\forall x . \Phi = \Phi [x \rightarrow \true] \land \Phi [x \rightarrow \false]$ and the \emph{existential} quantifier as $\exists x . \Phi = \Phi [x \rightarrow \true] \lor \Phi [x \rightarrow \false]$.

Let $X$ be a finite set of variables. Assignment $\mathcal{A}_{X}$ $:X \rightarrow \{\false, \true\}$ is a total mapping of variables defined on set $X$ to Boolean values.
Let $\Phi$ be a QBF. If there is an assignment $\mathcal{A}_{X}$ that evaluates formula $\Phi$ as $\true$, then $\Phi$ is $\mathit{satisfiable}$ (SAT).
A set $S \subseteq X$ is a \emph{model} of $\Phi$ if the assignment $\mathcal{A}_{X}$ maps $\true$ to every element of $S$. If there is no such assignment that evaluates the formula as $\true$, we say the formula is $\mathit{unsatisfiable}$ (UNSAT).

>>>>

\def\Mocc{M_{\mathrm{occur}}}
\def\Mevery{M_{\mathrm{every}}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\CA}{\ensuremath{\mathcal{A}}}
\newcommand{\CB}{\ensuremath{\mathcal{B}}}
\def\init{\iota}
\def\lang{L} % FIX this
\newcommand{\cM}{\mathcal{M}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\bscc}{\mathcal{B}}
\newcommand{\sched}{\mathfrak{s}}
\newcommand{\Act}{\mathit{Act}}
\newcommand{\NP}{\textsf{NP}}
\newcommand{\POLY}{\textsf{P}}
\newcommand{\class}{\ensuremath{\mathcal{C}}}
\newcommand{\finally}{\mathsf{F}}
\newcommand{\globally}{\mathsf{G}}
% \newcommand{\finally}{\mathbf{F}}
% \newcommand{\globally}{\mathbf{G}}

@@

\begin{abstract}
  The formalism called \emph{transition-based Emerson-Lei automata
    (TELA)} generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are marked with acceptance marks and
  its accepting formula is a positive Boolean combinations of terms
  saying that a particular mark has to be visited infinitely or
  finitely often in an accepting run. Algorithms processing these
  automata can be very sensitive to the number of different acceptance
  marks. We introduce two techniques reducing the number of acceptance
  marks in a TELA without altering its structure and the represented
  language. One technique represents a standard approach based on
  relations among acceptance marks in individual strongly connected
  components of the automaton. The other technique constructs
  \emph{quantified Boolean formula (QBF)} queries that ask a QBF
  solver for an acceptance condition with fewer acceptance marks and
  the placement of these marks in the automaton structure. Both
  techniques are implemented and experiments show that the number of
  acceptance marks in TELA produced by state-of-the-art tools
  ltl3tela, Rabinizer~4, and SPOT can be sometimes reduced. In the
  case of Rabinizer~4, we reduced the cummulative number of acceptance
  marks to less than one half.
\end{abstract}

@@

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity automata~\cite{renkin.20.atva}.

@@

The second technique is completely different. For a given TELA and
parameters $c,m$, it produces a \emph{quantified Boolean formula
  (QBF)} which is satisfiable if and only if there exists an automaton
with the same structure, $m$ acceptance marks, an acceptance formula
in disjunctive normal form with $c$ clauses, and the same set
of accepting runs. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of a
simpler formula such its satisfiability implies existence of an
automaton with the same structure, $m$ acceptance marks, and the same
set of accepting runs, but not vice versa.

@@

\begin{definition}[TELA]
  A \emph{transition-based Emerson-Lei automaton (TELA)} is a tuple
  $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$, where
  \begin{itemize}
  \item $Q$ is a finite set of \emph{states},
  \item $M$ is a finite set of \emph{acceptance marks},
  \item $\Sigma$ is a finite \emph{alphabet},
  \item $q_I\in Q$ is an \emph{initial state},
  \item $\delta \subseteq Q \times \Sigma \times 2^{M} \times Q$ is a
    \emph{transition relation}, and
  \item $\varphi$ is the \emph{acceptance condition} constructed
    according to the following abstract syntax equation, where $m$
    ranges over $M$:
    $$\varphi ::= \true \mid \false \mid \Inf(m) \mid \Fin(m) \mid \left( \varphi \land \varphi \right) \mid \left( \varphi \lor \varphi \right).$$
\end{itemize}
\end{definition}

%A transition $(p,a,M',q)\in\delta$ is a transition leading from state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$
A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $M(\pi)$ denote set of all acceptance marks that occur
in run $\pi$ in infinitely many transitions. Run $\pi$ is
\emph{accepting} iff it satisfies the formula $\varphi$, where $\pi$
satisfies $\Inf(m)$ iff $m\in M(\pi)$ and $\pi$ satisfies $\Fin(m)$
iff $m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the
set $L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of }\mA\textrm{ over }u\}$.
% such that $\mA$ has
% an accepting run over $u$.

An acceptance formula $\varphi$ is in \emph{disjunctive normal formal
  (DNF)} if it is a disjunction of clauses, where each clause is a
conjunction of some terms of the form $\Fin(m)$ or $\Inf(m)$.  Formula
$\false$ corresponds to the disjunction of zero clauses and formula
$\true$ is the clause with zero terms. Each acceptance formula can be
transformed into an equivalent formula in DNF.

For each mark $m\in M$, let $\delta_m$ denotes the set of all
transitions marked with $m$, i.e.,
$\delta_m=\{(p,a,M',q)\in\delta\mid m\in M'\}$. Further, a \emph{path}
from a state $p$ to a state $q$ is a finite sequence of adjacent
transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given an
SCC $S$, $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denotes the set of all transitions between states in $S$.

We often use acceptance marks $\tacc{1}{1},\tacc{2}{2},\ldots\in M$
and we write $\Inf\tacc{4}{j}$ and $\Fin\tacc{8}{k}$ instead of
$\Inf(\tacc{4}{j})$ and $\Fin(\tacc{8}{k})$, respectively. In our
graphical representation of automata,
\begin{tikzpicture}[smallautomaton]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.

% Finaly, \emph{loop} is a path from a state $p$ to the same state.

\bigskip

\emph{Quantified Boolean formulas (QBFs)} are Boolean formulas
extended with universal or existential quantification over
propositional variables. We assume that subformulas of the form
$\forall x.\psi$ and $\exists x.\psi$ do not contain another
quantification of variable $x$ inside $\psi$. The semantics of
$\forall x.\psi$ and $\exists x.\psi$ is given by equivalences
\[
  \begin{array}{rcl}
    \forall x.\psi &~\equiv~& \psi[x\rightarrow\true]~\wedge~\psi[x\rightarrow\false]\\
    \exists x.\psi &~\equiv~& \psi[x\rightarrow\true]~\vee~   \psi[x\rightarrow\false]
  \end{array}
\]
where $\psi[x\rightarrow \rho]$ denotes the formula $\psi$ with all
occurrences of $x$ simultaneously replaced by $\rho$. The equivalences
imply that each QBF can be transformed into an equivalent Boolean
formula. However, the size of this Boolean formula can be exponential
in the size of the original QBF. Let $V$ be the set of all
propositional variables. A mapping $\mu:V\rightarrow\{\true,\false\}$
is a \emph{model} of a QBF formula $\varphi$ iff it is a satisfying
assignment of the equivalent Boolean formula. A QBF formula is
\emph{satisfiable} iff it has a model.

A QBF $\varphi$ is in the \emph{prenex normal form} if it has the form
$\varphi = Q_{1}x_{1}...Q_{n}x_{n} . \psi$, where
$Q_1,Q_2,\ldots,Q_n \in \{\forall, \exists\}$ and $\psi$ is a Boolean
formula.
